---
title: "Exploratory Analysis for TB Report Card"
author: "Tampa Bay Estuary Program"
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
    theme: cosmo
    code-fold: true
    code-tools:
      source: https://github.com/tbep-tech/tampa-bay-report-card/blob/main/docs/exploratory.qmd
toc: true
lightbox: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tbeptools)
library(tidyverse)
library(here)
library(sf)
library(plotly)
library(jsonlite)
source(here("R", "funcs.R"))

load(file = here('data/sgsegest.RData'))
```

Indicators:
* Water Quality
    * chla, water clarity
    * loading
    * tidal creeks
    * FIBs
* Sediment
    * TBBI/AMBI
    * PEL/TEL
* Fish/Wildlife
    * TBNI
    * Nonnatives
* Habitat
    * Seagrass
    * LULC change/habitat report card

## Water Quality

```{r}
# wq
avedat <- anlz_avedat(epcdata)
wqtarg <- avedat |>
  anlz_attain() |>
  separate(chl_la, into = c("chl", "la"), sep = "_") |> 
  mutate(across(c(chl, la), as.numeric)) |>
  mutate(
    totsum = chl + la,
    outcome = 1 - scales::rescale(totsum, to = c(0, 1))
  )
wqthrs <- avedat$ann |>
    dplyr::left_join(targets, by = 'bay_segment') |>
    dplyr::mutate(
      bay_segment = factor(bay_segment, levels = c('OTB', 'HB', 'MTB', 'LTB')),
      outcome = dplyr::case_when(
        var == 'mean_chla' & val < chla_thresh ~ 1,
        var == 'mean_chla' & val >= chla_thresh ~ 0,
        var == 'mean_la' & val < la_thresh ~ 1,
        var == 'mean_la' & val >= la_thresh ~ 0
      )
    ) |> 
    select(yr, bay_segment, var, outcome)

# loading
totanndat <- rdataload("https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData")

segcrsswk <- c(
  'Old Tampa Bay' = 'OTB',
  'Hillsborough Bay' = 'HB',
  'Middle Tampa Bay' = 'MTB',
  'Lower Tampa Bay' = 'LTB'
)
tntarg <- tibble(
  bay_segment = factor(segcrsswk, levels = segcrsswk),
  tn_load_thresh = c(486, 1451, 799, 349),
  tnhy_thresh = c(1.08, 1.62, 1.24, 0.97)
)
tndat <- totanndat |>
  filter(bay_segment %in% names(segcrsswk)) |>
  mutate(bay_segment = factor(segcrsswk[bay_segment], levels = segcrsswk)) |>
  select(year, bay_segment, tn_load, hy_load, tnhy) |>
  left_join(tntarg, by = 'bay_segment') |>
  mutate(
    outcomeld = as.integer(tn_load > tn_load_thresh),
    outcomehy = as.integer(tnhy > tnhy_thresh)
  )

# tidal creeks
trnds <- tibble(
    yrs = 1975:2024,
  ) |> 
  group_nest(yrs) |> 
  mutate(
    data = purrr::map(yrs, function(x){
      # cat(x, '\t')
      anlz_tdlcrk(tidalcreeks, iwrraw, yr = x)
    }) 
  )

# assign each tidal creek to a bay segment based on the tbsegdetail
# subwatershed it overlaps, using length of overlap to pick a segment for
# creeks that span more than one subwatershed
tcseg <- tidalcreeks |>
  st_transform(st_crs(tbsegshed)) |>
  select(id) |>
  st_intersection(tbsegshed |> select(bay_segment)) |>
  mutate(complen = st_length(geometry)) |>
  st_drop_geometry() |>
  group_by(id) |>
  slice_max(complen, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(id, bay_segment)

tctrnds <- trnds |>
  unnest('data') |>
  inner_join(tcseg, by = 'id')

# condense the categorical creek scores into a single 0-1 value per bay
# segment and year (1 = best); creeks with 'No Data' are dropped from the
# calculation, and each creek record is weighted by its physical length so
# longer creek segments contribute proportionally more to the segment score
crkwts <- c(
  'Prioritize'  = 0,
  'Investigate' = 1/3,
  'Caution'     = 2/3,
  'Monitor'     = 1
)

tcscr <- tctrnds |>
  left_join(
    tidalcreeks |> st_drop_geometry() |> select(id, Creek_Length_m),
    by = 'id'
  ) |>
  filter(score %in% names(crkwts)) |>
  mutate(scoreval = crkwts[score]) |>
  group_by(bay_segment, yrs) |>
  summarise(
    outcome = weighted.mean(scoreval, w = Creek_Length_m, na.rm = TRUE),
    n_assessed = dplyr::n(),
    .groups = 'drop'
  )

# fibs
fibscr <- anlz_fibmatrix(enterodata, bay_segment = c('OTB', 'HB', 'MTB', 'LTB', 'BCB', 'MR')) |>
  mutate(
    outcome = case_when(
      cat == 'E' ~ 0,
      cat == 'D' ~ 1/4,
      cat == 'C' ~ 2/4,
      cat == 'B' ~ 3/4,
      cat == 'A' ~ 1
    )
  )
```

```{r}
# combine all water quality indicators into a single long-format object of
# bay_segment, year, indicator, and a 0-1 outcome (1 = best). coverage of
# bay segments and years differs by indicator, so this stacks the
# indicators (bind_rows) rather than joining them - a segment/year missing
# from an indicator's source data simply has no row for it, rather than an
# NA-filled one. tn_load/tnhy_load are flipped (1 - outcomeld/outcomehy) so
# 1 = attains target, matching the "1 = best" direction of every other
# indicator here
wqindic <- bind_rows(
  wqtarg |>
    transmute(
      bay_segment = as.character(bay_segment),
      yr = as.integer(yr),
      indicator = 'wq_attain',
      outcome
    ),
  wqthrs |>
    filter(var %in% c('mean_chla', 'mean_la')) |>
    transmute(
      bay_segment = as.character(bay_segment),
      yr = as.integer(yr),
      indicator = recode(var, mean_chla = 'chla_thresh', mean_la = 'la_thresh'),
      outcome
    ),
  tndat |>
    transmute(
      bay_segment = as.character(bay_segment),
      yr = as.integer(year),
      indicator = 'tn_load',
      outcome = 1 - outcomeld
    ),
  tndat |>
    transmute(
      bay_segment = as.character(bay_segment),
      yr = as.integer(year),
      indicator = 'tnhy_load',
      outcome = 1 - outcomehy
    ),
  tcscr |>
    transmute(
      bay_segment = as.character(bay_segment),
      yr = as.integer(yrs),
      indicator = 'tidal_creeks',
      outcome
    ),
  fibscr |>
    transmute(
      bay_segment = as.character(grp),
      yr = as.integer(yr),
      indicator = 'fib',
      outcome
    )
) |>
  filter(!is.na(outcome))
```

```{r}
# overall outcome by year and bay segment, averaged across whatever
# indicators have data for that segment/year. Restricted to OTB, HB, MTB,
# LTB (the four segments with the most complete indicator coverage) and to
# 2000 onward - by 2000, wq_attain/chla_thresh/la_thresh (from 1972-1974)
# and tn_load/tnhy_load (from 1985) are already available for all four
# segments every year, and fib (starting 2002/2003) and tidal_creeks'
# coverage of LTB (starting ~2005) both phase in shortly after. n_indicator
# tracks how many of the (up to 7) indicators fed each average, since a few
# early LTB years only have 5
wqoverall <- wqindic |>
  filter(bay_segment %in% c('OTB', 'HB', 'MTB', 'LTB'), yr >= 2000) |>
  group_by(bay_segment, yr) |>
  summarise(
    outcome = mean(outcome, na.rm = TRUE),
    n_indicator = dplyr::n(),
    .groups = 'drop'
  )
```

```{r}
# interactive "flower" plot for a single year: one polar bar (coxcomb)
# chart per bay segment, arranged 2x2. Each equal-angle petal is a
# component indicator (length/color = its 0-1 outcome) and the colored
# center hole is that segment's overall wqoverall score - the petals give
# context for what's driving the number in the middle. Any indicator
# missing for a segment/year (none in 2025) renders as a short gray "No
# data" petal rather than breaking the equal-angle layout
plotyr <- 2024

indic_labs <- c(
  wq_attain    = 'WQ Attain',
  chla_thresh  = 'Chl-a',
  la_thresh    = 'Light',
  tn_load      = 'TN Load',
  tnhy_load    = 'Hydro',
  tidal_creeks = 'Creeks',
  fib          = 'FIB'
)

segorder <- c('OTB', 'HB', 'MTB', 'LTB')
seglabs <- c(
  OTB = 'Old Tampa Bay',
  HB  = 'Hillsborough Bay',
  MTB = 'Middle Tampa Bay',
  LTB = 'Lower Tampa Bay'
)

# red -> orange -> yellow -> green, following the status palette used
# elsewhere for condition/attainment (worst to best)
colfun <- scales::col_numeric(
  palette = c('#d03b3b', '#ec835a', '#fab219', '#0ca30c'),
  domain  = c(0, 1)
)
nacolor <- '#c3c2b7'

petalsyr <- wqindic |>
  filter(yr == plotyr, bay_segment %in% segorder) |>
  complete(bay_segment = segorder, indicator = names(indic_labs)) |>
  mutate(
    bay_segment   = factor(bay_segment, levels = segorder),
    indicator_lab = factor(indic_labs[as.character(indicator)], levels = unname(indic_labs)),
    r_plot        = ifelse(is.na(outcome), 1, outcome),
    fillcolor     = ifelse(is.na(outcome), nacolor, colfun(outcome)),
    opacity       = ifelse(is.na(outcome), 0.25, 0.92),
    hovertext     = paste0(
      seglabs[as.character(bay_segment)], '<br>',
      indicator_lab, ': ',
      ifelse(is.na(outcome), 'No data', scales::percent(outcome, accuracy = 1))
    )
  )

centersyr <- wqoverall |>
  filter(yr == plotyr, bay_segment %in% segorder) |>
  mutate(bay_segment = factor(bay_segment, levels = segorder))

# explicit 2x2 domains (rather than plotly::subplot()) so the center hole
# size/position can be computed exactly for the score annotation below
domains <- list(
  OTB = list(x = c(0.04, 0.46), y = c(0.55, 1.00)),
  HB  = list(x = c(0.54, 0.96), y = c(0.55, 1.00)),
  MTB = list(x = c(0.04, 0.46), y = c(0.00, 0.45)),
  LTB = list(x = c(0.54, 0.96), y = c(0.00, 0.45))
)
polarnames <- c(OTB = 'polar', HB = 'polar2', MTB = 'polar3', LTB = 'polar4')
holefrac <- 0.34
figw <- 900
figh <- 840

fig <- plot_ly(width = figw, height = figh)
for (seg in segorder) {
  dd <- petalsyr |> filter(bay_segment == seg) |> arrange(indicator)
  fig <- fig |> add_trace(
    data       = dd,
    type       = 'barpolar',
    r          = ~r_plot,
    theta      = ~indicator_lab,
    marker     = list(color = ~fillcolor, opacity = ~opacity, line = list(color = 'white', width = 1)),
    subplot    = polarnames[[seg]],
    text       = ~hovertext,
    hoverinfo  = 'text',
    showlegend = FALSE
  )
}

polar_layout <- function(dom) list(
  domain      = dom,
  hole        = holefrac,
  bgcolor     = 'rgba(0,0,0,0)',
  radialaxis  = list(range = c(0, 1), showticklabels = FALSE, ticks = '', showline = FALSE, gridcolor = '#e1e0d9'),
  angularaxis = list(rotation = 90, direction = 'clockwise', tickfont = list(size = 9, color = '#52514e'), gridcolor = '#e1e0d9')
)

shapes <- list()
annos  <- list()
for (seg in segorder) {
  dom      <- domains[[seg]]
  cx       <- mean(dom$x)
  cy       <- mean(dom$y)
  qw_px    <- (dom$x[2] - dom$x[1]) * figw
  qh_px    <- (dom$y[2] - dom$y[1]) * figh
  r_px     <- min(qw_px, qh_px) / 2 * holefrac
  rx       <- r_px / figw
  ry       <- r_px / figh
  scoreval <- centersyr |> filter(bay_segment == seg) |> pull(outcome)
  ccol     <- colfun(scoreval)

  shapes[[length(shapes) + 1]] <- list(
    type = 'circle', xref = 'paper', yref = 'paper',
    x0 = cx - rx, x1 = cx + rx, y0 = cy - ry, y1 = cy + ry,
    fillcolor = ccol, line = list(color = 'white', width = 2)
  )
  annos[[length(annos) + 1]] <- list(
    x = cx, y = cy, xref = 'paper', yref = 'paper', xanchor = 'center', yanchor = 'middle',
    text = paste0('<b>', scales::percent(scoreval, accuracy = 1), '</b>'),
    showarrow = FALSE, font = list(size = 18, color = '#ffffff')
  )
  annos[[length(annos) + 1]] <- list(
    x = cx, y = dom$y[2] + 0.015, xref = 'paper', yref = 'paper', xanchor = 'center', yanchor = 'bottom',
    text = paste0('<b>', seglabs[[seg]], '</b>'),
    showarrow = FALSE, font = list(size = 13, color = '#0b0b0b')
  )
}

fig |>
  layout(
    title = list(text = paste('Bay Segment Indicator Scores —', plotyr), x = 0.5, font = list(size = 18)),
    showlegend = FALSE,
    polar  = polar_layout(domains$OTB),
    polar2 = polar_layout(domains$HB),
    polar3 = polar_layout(domains$MTB),
    polar4 = polar_layout(domains$LTB),
    shapes = shapes,
    annotations = annos,
    margin = list(t = 90, b = 20, l = 40, r = 40),
    paper_bgcolor = '#fcfcfb'
  )
```

## Sediment

```{r}
peltel <- tibble(
    yr = 1993:2024,
  ) |> 
  group_nest(yrs) |> 
  mutate(
    data = purrr::map(yr, function(x){
      anlz_sedimentpelave(sedimentdata, yrrng = x)
    }) 
  ) |> 
  unnest('data') |> 
  mutate(
    grd = cut(ave, breaks = c(-Inf, 0.00756, 0.02052, 0.08567, 0.28026, Inf), labels = c('A', 'B', 'C', 'D', 'F')), 
    outcome = case_when(
      grd == 'A' ~ 1,
      grd == 'B' ~ 0.75,
      grd == 'C' ~ 0.5,
      grd == 'D' ~ 0.25,
      grd == 'F' ~ 0
    )
  ) |> 
  select(yr, bay_segment, ave, grd, outcome)
tbbi <- benthicdata |> 
  anlz_tbbiscr() |> 
  anlz_tbbimed() |> 
  filter(!grepl('!All', bay_segment)) |> 
  mutate(
    outcome = case_when(
      TBBICat == 'Poor' ~ 0,
      TBBICat == 'Fair' ~ 0.5,
      TBBICat == 'Good' ~ 1
    )
  ) |> 
  select(yr, bay_segment, TBBICat, outcome)
```

## Fish/Wildlife


```{r}
tbni <- fimdata |> 
  anlz_tbniscr() |> 
  anlz_tbniave() |> 
  mutate(
    outcome = scales::rescale(Segment_TBNI, to = c(0, 1), from = c(0, 100))
  )

```

### Nonnatives

Ports the non-native species "report card" metrics from the
[tbep-invasives](https://github.com/tbep-tech) Python pipeline
(`src/tbep_invasives/steps/report_cards.py`) to R: non-native species
abundance (observations per unit area per year) and richness (unique
species per unit area per year) by bay segment, converted to percentiles.
Numeric output only (no plot) - two tidy year x bay-segment data frames,
`nonnative_abundance` and `nonnative_richness`.

```{r}
# raw occurrence data: USGS NAS API (9 HUC8 codes covering the Tampa Bay
# area, from tbep-invasives/config/settings.yaml) + the Tampa Bay FIM
# non-native dataset (an RData file - note the object stored inside is
# named `inv`, not derived from the URL basename, so rdataload() doesn't
# apply here)
huc8_list <- c(
  '03100101', '03100201', '03100202', '03100203',
  '03100204', '03100205', '03100206', '03100207', '03100208'
)

fetch_nas_huc8 <- function(huc8) {
  url <- paste0('https://nas.er.usgs.gov/api/v2/occurrence/search?huc8=', huc8)
  res <- jsonlite::fromJSON(url, simplifyDataFrame = TRUE)
  rows <- res$results
  if (is.null(rows) || nrow(rows) == 0) return(NULL)
  tibble(
    scientificName = rows$scientificName,
    commonName     = rows$commonName,
    group          = rows$group,
    year           = as.integer(rows$year),
    lon            = as.numeric(rows$decimalLongitude),
    lat            = as.numeric(rows$decimalLatitude)
  )
}

nas_raw <- huc8_list |>
  purrr::map(fetch_nas_huc8) |>
  bind_rows() |>
  mutate(source = 'NAS')

fim_tf <- tempfile(fileext = '.RData')
download.file(
  'https://github.com/kflahertywalia/tb_fim_nonnatives/raw/refs/heads/main/Output/tb_fim_inv.RData',
  fim_tf, mode = 'wb', quiet = TRUE
)
load(fim_tf)  # loads an object named `inv`

fim_raw <- inv |>
  transmute(
    scientificName = Scientificname,
    commonName     = Commonname,
    group          = recode(Taxa_Type, Fish = 'Fishes', Turtle = 'Reptiles'),
    year           = as.integer(substr(Reference, 4, 7)),
    lon            = as.numeric(Longitude),
    lat            = as.numeric(Latitude),
    source         = 'FIM'
  )

# group name cleanup (matches download_invasives.py's group_name_changes),
# keep only the same 7 taxonomic groups Python keeps
group_name_changes <- c(
  'Amphibians-Frogs'    = 'Amphibians',
  'Crustaceans-Crabs'   = 'Crustaceans',
  'Marine Fishes'       = 'Fishes',
  'Mollusks-Bivalves'   = 'Mollusks',
  'Mollusks-Gastropods' = 'Mollusks',
  'Reptiles-Lizards'    = 'Reptiles',
  'Reptiles-Snakes'     = 'Reptiles',
  'Reptiles-Turtles'    = 'Reptiles'
)
keep_groups <- c('Amphibians', 'Crustaceans', 'Fishes', 'Mammals', 'Mollusks', 'Plants', 'Reptiles')

nonnative_obs <- bind_rows(nas_raw, fim_raw) |>
  mutate(group = ifelse(group %in% names(group_name_changes), group_name_changes[group], group)) |>
  filter(group %in% keep_groups, year >= 2000, !is.na(lon), !is.na(lat))
```

```{r}
# bay segment / AOI boundaries: shapefiles copied from the tbep-invasives
# repo (via data-raw/, converted to RData by R/dat_proc.R) so segment
# grouping ("Remainder Lower Tampa Bay" = Boca Ciega Bay + Manatee River +
# Terra Ceia Bay) and areas match its report cards exactly
load(here('data/aoi_shp.RData'))
load(here('data/bayseg_shp.RData'))

nonnative_pts <- nonnative_obs |>
  st_as_sf(coords = c('lon', 'lat'), crs = 4326, remove = FALSE) |>
  st_filter(aoi_shp, .predicate = st_intersects) |>
  st_join(bayseg_shp['BAY_SEG_GP'], join = st_within) |>
  rename(baySegment = BAY_SEG_GP) |>
  st_drop_geometry() |>
  filter(!is.na(baySegment))

# areas in sq mi, recomputed from geometry (EPSG:3086) to match
# report_cards.py's calc_watershed_area()/calc_aoi_area() rather than
# trusting the shapefile's static area_ac attribute (the two agree closely
# where checked)
bayseg_area_sqmi <- bayseg_shp |>
  st_transform(3086) |>
  group_by(BAY_SEG_GP) |>
  summarise(geometry = st_union(geometry), .groups = 'drop') |>
  mutate(area_sqmi = as.numeric(st_area(geometry)) / 4046.8564224 / 640) |>
  st_drop_geometry() |>
  deframe()

aoi_area_sqmi <- sum(as.numeric(st_area(st_transform(aoi_shp, 3086)))) / 4046.8564224 / 640
```

```{r}
# per-segment abundance/richness normalized by area, plus an "All" column
# normalized by total AOI area - mirrors per_watershed_abundance(),
# per_watershed_richness(), total_aoi_abundance(), total_aoi_richness()
bay_seg_order <- c('Old Tampa Bay', 'Hillsborough Bay', 'Middle Tampa Bay', 'Lower Tampa Bay', 'Remainder Lower Tampa Bay')

ws_abundance <- nonnative_pts |>
  count(year, baySegment) |>
  pivot_wider(names_from = baySegment, values_from = n, values_fill = 0) |>
  select(year, all_of(bay_seg_order))
ws_abundance[bay_seg_order] <- Map(`/`, ws_abundance[bay_seg_order], bayseg_area_sqmi[bay_seg_order])

ws_richness <- nonnative_pts |>
  distinct(year, baySegment, scientificName) |>
  count(year, baySegment) |>
  pivot_wider(names_from = baySegment, values_from = n, values_fill = 0) |>
  select(year, all_of(bay_seg_order))
ws_richness[bay_seg_order] <- Map(`/`, ws_richness[bay_seg_order], bayseg_area_sqmi[bay_seg_order])

ws_abundance$All <- (nonnative_pts |> count(year) |> pull(n)) / aoi_area_sqmi
ws_richness$All  <- (nonnative_pts |> distinct(year, scientificName) |> count(year) |> pull(n)) / aoi_area_sqmi

# percentile transform: a single mean/sd across every cell of the whole
# matrix (including "All"), matching as_percentile_by_mean_std() exactly
as_percentile_by_mean_sd <- function(df) {
  vals <- as.matrix(df[, setdiff(names(df), 'year')])
  mu <- mean(vals, na.rm = TRUE)
  sigma <- sd(vals, na.rm = TRUE)
  pctl <- if (sigma <= 0 || is.na(sigma)) matrix(50, nrow(vals), ncol(vals)) else pnorm((vals - mu) / sigma) * 100
  out <- as_tibble(pctl)
  names(out) <- colnames(vals)
  bind_cols(year = df$year, out)
}

seg_abbrev <- c(
  'Old Tampa Bay'             = 'OTB',
  'Hillsborough Bay'          = 'HB',
  'Middle Tampa Bay'          = 'MTB',
  'Lower Tampa Bay'           = 'LTB',
  'Remainder Lower Tampa Bay' = 'RALTB'
)

# long format, bay segments abbreviated, "All" dropped (it's only needed
# above to match the global mean/sd used in as_percentile_by_mean_sd()).
# outcome reverses the percentile to a 0-1 scale where 1 = best, since
# higher non-native abundance/richness is worse
to_nonnative_outcome <- function(df) {
  df |>
    select(-All) |>
    pivot_longer(-year, names_to = 'bay_segment', values_to = 'value') |>
    mutate(
      yr = as.integer(year),
      bay_segment = seg_abbrev[bay_segment],
      outcome = 1 - value / 100
    ) |>
    select(yr, bay_segment, value, outcome)
}

nonnative_abundance <- as_percentile_by_mean_sd(ws_abundance) |> to_nonnative_outcome()
nonnative_richness  <- as_percentile_by_mean_sd(ws_richness) |> to_nonnative_outcome()
```

## Habitat

```{r}
# seagrass transects
trnsct <- transect |> 
  anlz_transectocc() |> 
  anlz_transectave() |> 
  filter(bay_segment != 'All') |> 
  mutate(
    outcome = scales::rescale(foest, to = c(0, 1), from = c(0, 100))
  ) |> 
  select(bay_segment, yr, foest, outcome)

# seagrass coverage
segtrgs <- tibble(
  segment = factor(c(levels(sgsegest$segment))),
  trgs = 1000 * c(11.1, 1.751, 9.4, 7.4, 8.8, 1.1, 0.449)
)
cov <- sgsegest |> 
  left_join(segtrgs, by = c('segment')) |> 
  mutate(
    outcome = ifelse(acres >= trgs, 1, 0)
  ) |> 
  select(segment, yr = year, acres, outcome) |> 
  complete(segment, yr = c(min(yr):2025)) |> 
  fill(acres, .direction = 'down', .by = segment) |>
  fill(outcome, .direction = 'down', .by = segment)
```

## Notes

* Indices that use categories from a continuous scale, revert to continuous data to get outcome
* For TBNI, the scores range from 0-100 but the categories have breaks at 32 and 46, take this into consideration. 
* For outcomes that are binary based on threshold, maybe use a sigmoidal conversion
* Lots of redundancy in the water quality indicators
* How to incorporate land use change by bay segment? Is this even appropriate?